Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2020-12-15 analysis/paper/paper.Rmd
Declaraciones
Correcciones de Leandro

abstract: |
  The main mode of variability in the Southern Hemisphere extratropical circulation is the Southern Annular Mode (SAM); so called because of it's zonally symmetric ring-like shape. However, the SAM pattern actually contains noticeable deviations from zonal symmetry. We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices. In the troposphere, the Asymmetric SAM, is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure. The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies. The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the A\nobreakdash-SAM index. The positive trend of the SAM is present only in its asymmetric component. Despite this, the SAM is becoming more zonally asymmetric. The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric component.

@@

Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and references therein]. It is though that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013]. However, it’s not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.

The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet. The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM). But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions. For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018]. @fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.

@@

Our objective is, then, to describe the zonally asymmetric and symmetric components of the SAM variability. We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively. Their vertical structure and coherence, temporal variability and trends are consequently assesed. We then study the spatial patterns described by the variability exclusive to each index focusing on 50\ hPa as representing the stratosphere and 700\ hPa as representing the troposphere. Finally, the relationships of the SAM at 700\ hPa with temperature and precipitation anomalies are investigated.

@@

We computed density probability estimates using a gaussian kernel of optimal bindwidth according to @sheather1991.

@@

From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM series (see, for example, the positive events on late 1987). Figure \@ref(fig:cor-lev) shows these correlations along all levels considered, for zero and -1 lags. Values of zero-lag correlations between A\nobreakdash-SAM and S\nobreakdash-SAM are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev`\ hPa and then increase again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.

@@

By separating the SAM signal in its asymmetric and symmetric parts, notonly we can see that these trends are almost entirely due to the symmetric component (column b vs. column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer. In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`). In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.

@@

In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole. On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric. Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere

@@

```{r wave-amplitude, fig.cap = "(ref:wave-amplitude-caption)", fig.width=4, fig.height=3, fig.env = "figure*", cache = FALSE}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  scale_color_brewer("Wavenumber", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", position = "tr")  +
  no_grid 
```

The amplitude of the first zonal wave numbers at each latitude at 50\ hPa and 700\ hPa are shown in Figure \@ref(fig:wave-amplitude), where wave number zero represents the amplitude of the zonal mean. Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels. However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`). Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean. On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.

@@

```{r regr-air-season, fig.cap = "Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\\nobreakdash-SAM and S\\nobreakdash-SAM for the 1979 -- 2018 period. Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate. Gray areas in Antarctica have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
air_regr %>% 
  copy() %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), 
                    breaks = AnchorBreaks(0, 0.1, 0)) +
  # geom_contour_tanaka(aes(z = estimate),
  #                     breaks = AnchorBreaks(0, 0.1, 0)) +
  # geom_raster(aes(fill = estimate)) +
  # geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), fill = "gray50", alpha = 0.2) +
  geom_contour2(aes(z = p.value), breaks = c(0, 0.05), color = "black", size = 0.3) +
  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray") +
  geom_contour2(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), color = "gray50") +
  geom_qmap(~.x[lat <= 0]) +
  scale_x_longitude() +
  scale_y_latitude(labels = NULL, limits = c(NA, 0), 
                   minor_breaks = NULL) +
  scale_fill_divergent("2-meter air tempreature",  
                       limits = c(-0.6, .6), 
                       oob = scales::squish,
                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
  coord_polar() +
  facet_grid(season ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", tag_suffix = "") +
  no_grid +
  axis_labs_smol +
  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
```

@@

Regression of the SAM indices with seasonal mean precipitation and 700\ hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for Australasia and South America respectively. South Africa is not shown because no significant signal was detected there.

@@

In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006. The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1). Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007. A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.

@@

The anomalous circulation at 700\ hPa associated with S\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile. This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region [e.g. @garreaud2007]. On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North. This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009;@zamboni2010]. This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018]. There is a small area of significative positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.

@@

The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation. There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980;@watson2014;@zhang2020], so it is not unlikely that the SAM would be similarly affected. However, establishing this link would require further research. 
We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels. We show that these trends maximise at 100\ hPa, and are explained by the zonally symmetric component. We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index. We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM. This is in contrast to @fogt2012 who made the evaluation for the period 158 -- 2001. This discrepancy might be due either due to differneces in methodology or analysed period.

In the troposphere, the spatial patterns of geopotential associated with the S\nobreakdash-SAM index are much closer to being fully annular than the patterns associated with the SAM index. The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific. This pattern extends vertically in the troposphere but its maximum is located at 250\ hPa, which also could suggest that surface-based indices are not opitimal for capturing this variability. This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability. We found that the significant correlation that exists between the SAM index and the Oceanic Niño Index over this period is captured entirely by the A\nobreakdash-SAM index. This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the A\nobreakdash-SAM index could be a useful measure to further study that relationship.

@@

By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts. This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component [@arblaster2006].

@@

## Data statement

All data used in this paper is freely available from their respective sources. ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means). NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html). The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt

@@

indices_wide %>% 
  .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)] %>% 
  ggplot(aes(lag, lev)) +
  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
  geom_contour_tanaka(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
  geom_vline(xintercept = 0) +
  scale_fill_divergent(NULL,
                       guide = guide_colorstrip_bottom(20),
                       breaks = AnchorBreaks(0, 0.05),
                       limits = c(NA, .59)) +
  scale_x_continuous("Lag (months)\nSymmetric leads Asymmetric\t \tAsymmetric leads Symmetric", 
                     breaks = seq(-5, 5, 1), 
                     expand = c(0, 0)) +
  scale_y_level()
```

@@

>>>>

abstract: |
  The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extra-tropical circulation and it is so called because of its zonally symmetric ring-like shape. However, the SAM pattern actually contains noticeable deviations from zonal symmetry. Thus, the purpose of this study is to describe the zonally asymmetric and symmetric components of the SAM variability and their impacts. We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices: Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM). In the troposphere, the A-SAM is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure. On the other hand, the S-SAM is associated with negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies. The observed relationship between the El Niño Southern Oscillation and the SAM is fully explained by the A-SAM index. The positive trend of the SAM is present only in its symmetric component. Despite this, the SAM is becoming more zonally asymmetric. The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric component.

@@

Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and references therein]. It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013]. However, it’s not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.

The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet. The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM). But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions. For example, the SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018]. @fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.

@@

Our objective is, then, to describe the zonally asymmetric and symmetric components of the SAM variability. We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively. Their vertical structure and coherence, temporal variability and trends are consequently assessed. We then study the spatial patterns described by the variability exclusive to each index focusing on 50\ hPa as representing the stratosphere and 700\ hPa as representing the troposphere. Finally, the relationships of the SAM at 700\ hPa with temperature and precipitation anomalies are investigated.

@@

We computed density probability estimates using a Gaussian kernel of optimal bin width according to @sheather1991.

@@

From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM series (see, for example, the positive events on late 1987). Figure \@ref(fig:cor-lev) shows these correlations along all levels considered, for zero and -1 lags. Values of zero-lag correlations between A\nobreakdash-SAM and S\nobreakdash-SAM are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev`\ hPa and then increase again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index. (Correlations at lags -5 to 5 are shown in Figure \@ref(fig:A1).)

@@

By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs. column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer. In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`). In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.

@@

In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole. On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric. Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere

@@

```{r wave-amplitude, fig.cap = "(ref:wave-amplitude-caption)", fig.width=4, fig.height=3, fig.env = "figure*", cache = FALSE}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  scale_color_brewer("wavenumber", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", position = "tr")  +
  no_grid 
```

The amplitude of the first zonal wavenumbers at each latitude at 50\ hPa and 700\ hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero represents the amplitude of the zonal mean. Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels. However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`). Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean. On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.

@@

```{r regr-air-season, fig.cap = "Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\\nobreakdash-SAM and S\\nobreakdash-SAM for the 1979 -- 2018 period. Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate. Gray areas in Antarctica have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
air_regr %>% 
  copy() %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), 
                    breaks = AnchorBreaks(0, 0.1, 0)) +
  # geom_contour_tanaka(aes(z = estimate),
  #                     breaks = AnchorBreaks(0, 0.1, 0)) +
  # geom_raster(aes(fill = estimate)) +
  # geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), fill = "gray50", alpha = 0.2) +
  geom_contour2(aes(z = p.value), breaks = c(0, 0.05), color = "black", size = 0.3) +
  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray") +
  geom_contour2(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), color = "gray50") +
  geom_qmap(~.x[lat <= 0]) +
  scale_x_longitude() +
  scale_y_latitude(labels = NULL, limits = c(NA, 0), 
                   minor_breaks = NULL) +
  scale_fill_divergent("2-meter air temperature",  
                       limits = c(-0.6, .6), 
                       oob = scales::squish,
                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
  coord_polar() +
  facet_grid(season ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", tag_suffix = "") +
  no_grid +
  axis_labs_smol +
  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
```

@@

Regression of the SAM indices with seasonal mean precipitation and 700\ hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South America respectively. South Africa is not shown because no significant signal was detected there.

@@

In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the Southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006. The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1). Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007. A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.

@@

The anomalous circulation at 700\ hPa associated with S\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile. This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region [e.g. @garreaud2007]. On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North. This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009;@zamboni2010]. This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018]. There is a small area of significant positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.

@@

The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation. There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980;@watson2014;@zhang2020], so it is not unlikely that the SAM would be similarly affected. However, establishing this link would require further research.

We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels. We show that these trends maximise at 100\ hPa, and are explained by the zonally symmetric component. We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index. We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM. This is in contrast to @fogt2012 who made the evaluation for the period 1958 -- 2001. This discrepancy might be due either due to differences in methodology or analysed period.

In the troposphere, the spatial patterns of geopotential associated with the S\nobreakdash-SAM index are much closer to being fully annular than the patterns associated with the SAM index. The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific. This pattern extends vertically in the troposphere but its maximum is located at 250\ hPa, which also could suggest that surface-based indices are not optimal for capturing this variability. This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability. We found that the significant correlation that exists between the SAM index and the Oceanic Niño Index over this period is captured entirely by the A\nobreakdash-SAM index. This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the A\nobreakdash-SAM index could be a useful measure to further study that relationship.

@@

By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts. This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component [@arblaster2006].

@@

# Declarations {-}

## Funding {-}

The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.

## Conflicts of interest {-}

The authors declare that they have no conflict of interest.

## Availability of data and material {-}

All data used in this paper is freely available from their respective sources. ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means). NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sciences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html). The Oceanic Niño Index is available via NOAA's Climate Prediction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt

## Code availability {-}

@@

@@

indices_wide %>% 
  .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)] %>% 
  ggplot(aes(lag, lev)) +
  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
  geom_contour_tanaka(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
  geom_vline(xintercept = 0) +
  scale_fill_divergent(NULL,
                       guide = guide_colorstrip_bottom(20),
                       breaks = AnchorBreaks(0, 0.05),
                       limits = c(NA, .59)) +
  scale_x_continuous("Lag (months)\nA-SAM leads S-SAM              S-SAM leads A-SAM", 
                     breaks = seq(-5, 5, 1), 
                     expand = c(0, 0)) +
  scale_y_level()
```
